---
title: "Clustering"
author: "Ville Lehtonen"
date: "11/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working dir to the source file location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Clustering
Here, we perform clustering of our primary datasets (uncorrected and cell cycle corrected datasets). The aim is to see if cells from same samples cluster together and form distinct communities and whether the correction of cell cycle effect affects this clustering.

## 1. Load packages and data
### 1.1 Load packages
```{r load-packages, message=FALSE}
library(SingleCellExperiment)
library(scran)
library(igraph)
library(scater)
library(ggplot2)
library(pheatmap)
library(fossil)
library(cluster)
```

### 1.2 Load pre-calculated data
**The clusters are calculated using Puhti supercomputer and results are loaded in the code chunk below to speed up the process**
```{r primary-data-clustering-shortcut}
# Load primary dataset
denoised.sce <- readRDS("./../saved/R_objects/denoised_sce_clust")

# Load cluster assignments and SNN graphs
snng <- readRDS("./../saved/R_objects/snng")
snng.cc <- readRDS("./../saved/R_objects/snng_cc")
clust <- readRDS("./../saved/R_objects/clust")
clust.cc <- readRDS("./../saved/R_objects/clust_cc")

# Make the cc corrected data a separate object
denoised.sce.cc <- as(altExp(denoised.sce, "CC_corrected"), "SingleCellExperiment")
denoised.sce.cc$cluster.cc <- factor(clust.cc)
reducedDim(denoised.sce.cc, "UMAP_cc") <- reducedDim(denoised.sce, "UMAP_cc")
reducedDim(denoised.sce.cc, "TSNE_cc") <- reducedDim(denoised.sce, "TSNE_cc")
reducedDim(denoised.sce.cc, "force_cc") <- reducedDim(denoised.sce, "force_cc")
```


## 2. Perform clustering (optional)
**Perform clustering (in case no pre-calculated files are available)**
```{r primary-data-clustering, eval=FALSE}
# Load data
denoised.sce <- readRDS("./../saved/R_objects/denoised_sce")

# Get # of reduced dimensions for uncorrected & cc-corrected datasets
dimensions <- ncol(reducedDim(denoised.sce, "PCA"))
dimensions.cc <- ncol(reducedDim(denoised.sce, "PCA_cc"))
dimensions
dimensions.cc

# Build SNN graph with k=180
snng <- buildSNNGraph(denoised.sce, k = 180, use.dimred = "PCA") 

denoised.sce.cc <- as(altExp(denoised.sce, "CC_corrected"), "SingleCellExperiment")
reducedDim(denoised.sce.cc, "PCA_cc") <- reducedDim(denoised.sce, "PCA_cc")
snng.cc <- buildSNNGraph(denoised.sce.cc, k = 180, use.dimred = "PCA_cc")

# Perform clustering and assign cluster labels
clust <- igraph::cluster_walktrap(snng)$membership
clust.cc <- igraph::cluster_walktrap(snng.cc)$membership
table(clust)
table(clust.cc)
denoised.sce$cluster <- factor(clust)
denoised.sce$cluster.cc <- factor(clust.cc)

# Run UMAP, t-SNE, and force-directed layout
## Uncorrected dataset
set.seed(123)
reducedDim(denoised.sce, "force") <- igraph::layout_with_fr(snng)
denoised.sce <- runUMAP(denoised.sce, dimred = "PCA", n_dimred = dimensions)
denoised.sce <- runTSNE(denoised.sce, dimred = "PCA", n_dimred = dimensions)
## Cc corrected data
reducedDim(denoised.sce, "force_cc") <- igraph::layout_with_fr(snng.cc)
denoised.sce <- runTSNE(denoised.sce, altexp = "CC_corrected", exprs_values = "scaled_logcounts", dimred = "PCA_cc", name = "TSNE_cc")
denoised.sce.cc <- runUMAP(denoised.sce.cc, exprs_values = "scaled_logcounts", dimred = "PCA_cc", name = "UMAP_cc", n_neighbors = 180)

# Set reduced dims to cc corrected experiment and vice versa
reducedDim(denoised.sce, "UMAP_cc") <- reducedDim(denoised.sce.cc, "UMAP_cc")
reducedDim(denoised.sce.cc, "TSNE_cc") <- reducedDim(denoised.sce, "TSNE_cc")
reducedDim(denoised.sce.cc, "force_cc") <- reducedDim(denoised.sce, "force_cc")

# Save sce file
#saveRDS(denoised.sce, file = "./../saved/R_objects/denoised_sce_clust")
```


## 3. Visualizing clustering results: Force-directed layout, t-SNE, and UMAP plots
```{r primary-data-clustering-plots, message=FALSE, warning=FALSE}
# force directed layout plots
## uncorrected
p.force <- plotReducedDim(denoised.sce, 
                          colour_by = "cluster",
                          text_by = "sample",
                          dimred = "force",
                          point_alpha = 0.4,
                          point_size = 1) +
  ggtitle("Force-directed layout to visualize variation \nin single cell transcriptomes",
          subtitle = "Force-directed layout of primary data:\nNumber of nearest neighbours (k): 180,\ncells coloured by sample groups and clusters (1-5) shown in the plot") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold")) +
  scale_fill_discrete(type = c("#E41A1C", "#377EB8", "#4DAF4A", "#FFFF33", "#FF7F00"),
                      guide = guide_legend(title = "Cluster", override.aes = list(size = 5, alpha = 1)))

## cc corrected
p.force.cc <- plotReducedDim(denoised.sce.cc, 
                             colour_by = "cluster.cc",
                             text_by = "sample",
                             dimred = "force_cc",
                             point_alpha = 0.4,
                             point_size = 1) +
  ggtitle("Force-directed layout to visualize variation \nin single cell transcriptomes",
          subtitle = "Force-directed layout of cell cycle corrected primary data:\nNumber of nearest neighbours (k): 180, cells coloured by\nsample groups and clusters (1-5) shown in the plot") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold")) +
  scale_fill_discrete(type = c("#E41A1C", "#377EB8", "#4DAF4A", "#FFFF33", "#FF7F00"),
                      guide = guide_legend(title = "Cluster", override.aes = list(size = 5, alpha = 1)))

p.force
p.force.cc

# t-SNE plots
## Uncorrected
p.tsne <- plotReducedDim(denoised.sce, 
                         colour_by = "cluster",
                         text_by = "sample",
                         dimred = "TSNE",
                         point_alpha = 0.4,
                         point_size = 1) +
  ggtitle("t-SNE projection to visualize variation \nin single cell transcriptomes",
          subtitle = "t-SNE plot of primary data: Number of nearest neighbours (k): 180,\ncells coloured by sample groups and clusters (1-5) shown in the plot") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold")) +
  scale_fill_discrete(type = c("#E41A1C", "#377EB8", "#4DAF4A", "#FFFF33", "#FF7F00"),
                      guide = guide_legend(title = "Cluster", override.aes = list(size = 5, alpha = 1)))

## cc corrected
p.tsne.cc <- plotReducedDim(denoised.sce.cc, 
                            colour_by = "cluster.cc",
                            text_by = "sample",
                            dimred = "TSNE_cc",
                            point_alpha = 0.4,
                            point_size = 1) +
  ggtitle("t-SNE projection to visualize variation \nin single cell transcriptomes after cell\ncycle correction",
          subtitle = "t-SNE plot of cell cycle corrected primary data: Number of\nnearest neighbours (k): 180, cells coloured by sample groups\nand clusters (1-5) shown in the plot") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold")) +
  scale_fill_discrete(type = c("#E41A1C", "#377EB8", "#4DAF4A", "#FFFF33", "#FF7F00"),
                      guide = guide_legend(title = "Cluster", override.aes = list(size = 5, alpha = 1)))

p.tsne
p.tsne.cc

# UMAP plots
## Uncorrected
p.umap <- plotReducedDim(denoised.sce, 
                         colour_by = "cluster",
                         text_by = "sample",
                         dimred = "UMAP",
                         point_alpha = 0.4,
                         point_size = 1) +
  ggtitle("UMAP projection to visualize variation \nin single cell transcriptomes",
          subtitle = "UMAP plot of primary data: Number of nearest neighbours (k): 180,\ncells coloured by sample groups and clusters (1-5) shown in the plot") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold")) +
  scale_fill_discrete(type = c("#E41A1C", "#377EB8", "#4DAF4A", "#FFFF33", "#FF7F00"),
                      guide = guide_legend(title = "Cluster", override.aes = list(size = 5, alpha = 1)))


## CC-corrected
p.umap.cc <- plotReducedDim(denoised.sce.cc, 
                            colour_by = "cluster.cc",
                            text_by = "sample",
                            dimred = "UMAP_cc",
                            point_alpha = 0.4,
                            point_size = 1) +
  ggtitle("UMAP projection to visualize variation \nin single cell transcriptomes after cell\ncycle correction",
          subtitle = "UMAP plot of cell cycle corrected primary data: Number of\nnearest neighbours (k): 180, cells coloured by sample groups\nand clusters (1-5) shown in the plot") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 20, face = "bold")) +
  scale_fill_discrete(type = c("#E41A1C", "#377EB8", "#4DAF4A", "#FFFF33", "#FF7F00"),
                      guide = guide_legend(title = "Cluster", override.aes = list(size = 5, alpha = 1)))

p.umap
p.umap.cc
```


**Save UMAP and t-SNE plots in a pdf file**
```{r umap-tsne-into-pdf, eval=FALSE}
pdf("./../saved/figures/primary_data_umap_and_tsne_plots.pdf", width = 7, height = 7)
p.tsne
p.tsne.cc
p.umap
p.umap.cc
dev.off()
```



## 4. Clustering comparison
**Compare the uncorrected and cell cycle corrected clustering results by adjusted rand index (ARI) and graph modularity**
```{r primary-data-clustering-comparison}
# Calculate modularities of the graphs
mod <- modularity(snng, clust)
mod.cc <- modularity(snng.cc, clust.cc)
mod
mod.cc

# Calculate adjusted Rand index
ari <- adj.rand.index(clust, clust.cc)
ari
```


**Calculate the modularity of each cluster and visualize the results**
```{r plot-modularities}
# Plot the modularities of clusters
ratio <- clusterModularity(snng, clust, as.ratio = TRUE)
ratio.cc <- clusterModularity(snng.cc, clust.cc, as.ratio = TRUE)

pheatmap(log2(ratio + 1), cluster_cols = FALSE, cluster_rows = FALSE, color=colorRampPalette(c("white", "blue"))(100), main = "Primary data, k = 180")
pheatmap(log2(ratio.cc + 1), cluster_cols = FALSE, cluster_rows = FALSE, color=colorRampPalette(c("white", "blue"))(100), main = "Primary data (cc corrected), k = 180")
```


**Silhouette analysis: (Optional due to being computationally a heavy operation)**
```{r silhouette-analysis, eval=FALSE}
# Perform silhouette analysis
dist.data <- dist(reducedDim(denoised.sce, "PCA"))
dist.data.cc <- dist(reducedDim(denoised.sce, "PCA_cc"))
sil <- silhouette(clust, dist = dist.data)
sil.cc <- silhouette(clust.cc, dist = dist.data.cc)
plot(sil, main = "Primary data: silhouette plot")
plot(sil, main = "Primary data (cc corrected): silhouette plot")

# Silhouette with subset of cells
denoised.sce.f <- denoised.sce[, seq(1, 21808, by = 80)]
denoised.sce.f.cc <- denoised.sce[, seq(1, 21808, by = 80)]
clust.f <- clust[seq(1, 21808, by = 80)]
clust.f.cc <- clust.cc[seq(1, 21808, by = 80)]
dist.data.f <- dist(reducedDim(denoised.sce.f, "PCA"))
dist.data.f.cc <- dist(reducedDim(denoised.sce.f.cc, "PCA"))
sil.f <- silhouette(clust.f, dist = dist.data.f)
sil.f.cc <- silhouette(clust.f.cc, dist = dist.data.f.cc)

plot(sil.f, main = "Subset of primary data: silhouette plot")
plot(sil.f.cc, main = "Subset of primary data (cc corrected): silhouette plot")

# Plot a heatmap of the differences in clustering between original data and cc-corrected data
tab <- table(clust, clust.cc)
pheatmap(log10(tab+10), main = "Original data vs. CC corrected (k=180)", color=viridis::viridis(100))
```



