---
title: "DE_analysis"
author: "Ville Lehtonen"
date: "11/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working dir to the source file location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# Marker gene detection
This script performs DE analysis between the control and the cap samples and between the clusters. We use pairwise Wilcoxon Mann Whitney (WMW) test to test for DE and use the following thresholds: Log-fold-change >= Log2(1.5), and FDR <= 0.1 and AUCs of >= 0.7/0.8 for strong upregulation and AUC of 0.2-0.3 for strong downregulation.

The aim is to identify which genes drive separation between the clusters identifies by the **Clustering.Rmd** script.


## 1. Load packages and data
### 1.1 Load packages
```{r load-packages, message=FALSE}
library(SingleCellExperiment)
library(scran)
library(pheatmap)
library(Seurat)
library(plyr)
```

### 1.2 Load pre-calculated data
**Single Cell Experiments**
```{r load-objects}
# Load pre-processed sce objects
denoised.sce <- readRDS("./../saved/R_objects/denoised_sce_clust")

# Generate the CC corrected data
denoised.sce.cc <- as(altExp(denoised.sce, "CC_corrected"), "SingleCellExperiment")
denoised.sce.cc$cluster.cc <- denoised.sce$cluster.cc
```

## 1. Primary data marker genes
### 1.1 Pairwise comparisons with t-test
**Markers by sample: Control**
```{r de-genes-primary-data-samples, eval=FALSE}
# Find markers by sample
markers.primary <- findMarkers(denoised.sce, 
                               groups = denoised.sce$sample,
                               lfc = log2(1.5))

# Choose markers for control sample
control.markers <- markers.primary[["control"]]

# Save marker information to .csv file
write.csv(control.markers, file = "./../saved/tables/control_marker_genes.csv")

# Subset top 6 genes (ranked by p -value) from each pairwise comparison involving control sample
best.set <- control.markers[control.markers$Top <= 6,]

# Extract the marker effect sizes
logFCs <- getMarkerEffects(best.set)

# Plot results in a heatmap
pheatmap(logFCs, breaks=seq(-5, 5, length.out=101), show_rownames = TRUE, show_colnames = TRUE, main = "Marker genes (t-test): Control vs. other samples")
```

**Markers by cluster**
```{r de-genes-primary-data-clusters, eval=FALSE}
# Find markers by sample
markers.primary.clust <- findMarkers(denoised.sce, 
                               groups = denoised.sce$cluster,
                               lfc = log2(1.5))

# Perform marker gene identification for all clusters
for(cluster in unique(denoised.sce$cluster)) {
  # Choose marker for cluster x
  markers <- markers.primary.clust[[cluster]]
  # Save marker information to .csv file
  fpath <- paste0("./../saved/tables/marker_genes_cluster_", cluster, ".csv")
  write.csv(markers, file = fpath)
  # Subset top 6 genes (ranked by p -value) from each pairwise comparison involving the cluster x
  best.set <- markers[markers$Top <= 6,]
  # Extract the marker effect sizes
  logFCs <- getMarkerEffects(best.set)
  # Plot results in a heatmap
  pheatmap(logFCs, breaks=seq(-5, 5, length.out=101), show_rownames = TRUE, show_colnames = TRUE, main = paste("Marker genes (t-test): cluster:", cluster))
}
```


**Markers by cell cycle corrected clusters**
```{r de-genes-primary-data-clusters-cc, eval=FALSE}
# Find markers by sample
markers.primary.clust.cc <- findMarkers(denoised.sce, 
                               groups = denoised.sce$cluster.cc,
                               lfc = log2(1.5))

# Perform marker gene identification for all clusters
for(cluster in unique(denoised.sce$cluster.cc)) {
  # Choose marker for cluster x
  markers <- markers.primary.clust.cc[[cluster]]
  # Save marker information to .csv file
  fpath <- paste0("./../saved/tables/marker_genes_cluster_", cluster, "_cc.csv")
  write.csv(markers, file = fpath)
  # Subset top 6 genes (ranked by p -value) from each pairwise comparison involving the cluster x
  best.set <- markers[markers$Top <= 6,]
  # Extract the marker effect sizes
  logFCs <- getMarkerEffects(best.set)
  # Plot results in a heatmap
  pheatmap(logFCs, breaks=seq(-5, 5, length.out=101), show_rownames = TRUE, show_colnames = TRUE, main = paste("Marker genes (t-test): cc corected cluster:", cluster))
}
```


### 1.2 Pairwise comparisons with Wilcoxon rank sum (Wilcoxon-Mann-Whitney, wmw)  test
**Markers by sample: Control & Cap50 samples**
```{r de-genes-primary-data-samples-wmw}
# Find markers by sample
markers.primary.wmw <- findMarkers(denoised.sce, 
                                   groups = denoised.sce$sample,
                                   lfc = log2(1.5),
                                   test = "wilcox")

# Choose markers for control & cap50 samples
control.markers.wmw <- markers.primary.wmw[["control"]]
cap50.markers.wmw <- markers.primary.wmw[["cap50"]]

# Flip the direction of up / down regulation
control.markers.wmw <- data.frame(as.data.frame(control.markers.wmw)[1:3], apply(control.markers.wmw[4:8], 2, function(x) {abs(1-x)}))
cap50.markers.wmw <- data.frame(as.data.frame(cap50.markers.wmw)[1:3], apply(cap50.markers.wmw[4:8], 2, function(x) {abs(1-x)}))

# Save marker information to .csv file
write.csv(control.markers.wmw, file = "./../saved/tables/control_marker_genes_wmw.csv")
write.csv(cap50.markers.wmw, file = "./../saved/tables/cap50_marker_genes_wmw.csv")

# Print and save DE genes with FDR and summary.AUC thresholds
## Control
de.genes.control <- control.markers.wmw[(control.markers.wmw$FDR <= 0.1) & (control.markers.wmw$summary.AUC >= 0.7 |control.markers.wmw$summary.AUC <= 0.3), ]
print(rownames(de.genes.control))
write.csv(de.genes.control, file = "./../saved/tables/control_marker_genes_wmw_filtered.csv")

strong.de.genes.control <- control.markers.wmw[(control.markers.wmw$FDR <= 0.1) & (control.markers.wmw$summary.AUC >= 0.8 |control.markers.wmw$summary.AUC <= 0.2), ]
print(rownames(strong.de.genes.control))
write.csv(strong.de.genes.control, file = "./../saved/tables/control_marker_genes_wmw_filtered_strong.csv")

## Cap 50
de.genes.cap50 <- cap50.markers.wmw[(cap50.markers.wmw$FDR <= 0.1) & (cap50.markers.wmw$summary.AUC >= 0.7 |cap50.markers.wmw$summary.AUC <= 0.3), ]
print(rownames(de.genes.cap50))
write.csv(de.genes.cap50, file = "./../saved/tables/cap50_marker_genes_wmw_filtered.csv")

strong.de.genes.cap50 <- cap50.markers.wmw[(cap50.markers.wmw$FDR <= 0.1) & (cap50.markers.wmw$summary.AUC >= 0.8 |cap50.markers.wmw$summary.AUC <= 0.2), ]
print(rownames(strong.de.genes.cap50))
write.csv(strong.de.genes.cap50, file = "./../saved/tables/cap50_marker_genes_wmw_filtered_strong.csv")

# Subset top 6 genes (ranked by p -value) from each pairwise comparison involving control / cap50 sample
best.set.control <- control.markers.wmw[control.markers.wmw$Top <= 6,]
best.set.cap50 <- cap50.markers.wmw[cap50.markers.wmw$Top <= 6,]

# Extract the area under curves
## Control
AUCs.best.control <- getMarkerEffects(best.set.control, prefix="AUC")
AUCs.de.control <- getMarkerEffects(de.genes.control, prefix="AUC")
AUCs.de.strong.control <- getMarkerEffects(strong.de.genes.control, prefix="AUC")

AUCs.de.control <- AUCs.de.control[, c("cap50", "cap50_r4h", "cap50_r8h", "cap50_r16h")]
AUCs.de.strong.control <- AUCs.de.strong.control[, c("cap50", "cap50_r4h", "cap50_r8h", "cap50_r16h")]

## Cap50
AUCs.best.cap50 <- getMarkerEffects(best.set.cap50, prefix="AUC")
AUCs.de.cap50 <- getMarkerEffects(de.genes.cap50, prefix="AUC")
AUCs.de.strong.cap50 <- getMarkerEffects(strong.de.genes.cap50, prefix="AUC")

AUCs.de.cap50 <- AUCs.de.cap50[, c("cap50_r4h", "cap50_r8h", "cap50_r16h", "control")]
AUCs.de.strong.cap50 <- AUCs.de.strong.cap50[, c("cap50_r4h", "cap50_r8h", "cap50_r16h", "control")]


# Plot results in a heatmap
## Control
pheatmap(AUCs.best.control, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21), main = "Marker genes (wmw): Control vs. other samples (Top 6)")
pheatmap(AUCs.de.strong.control, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21), main = "Marker genes (wmw): Control vs. other samples (FDR <= 0.2 & AUC >= 0.8 | AUC <= 0.2)", fontsize_row = 3)

## Cap50
pheatmap(AUCs.best.cap50, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21), main = "Marker genes (wmw): cap50 vs. other samples (Top 6)")
pheatmap(AUCs.de.strong.cap50, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21), main = "Marker genes (wmw): cap50 vs. other samples (FDR <= 0.2 & AUC >= 0.8 | AUC <= 0.2)", fontsize_row = 3)


# Count the number of upregulated and downregulated genes
n_upregulated = nrow(de.genes.control[de.genes.control$summary.AUC >= 0.7, ])
n_downregulated = nrow(de.genes.control[de.genes.control$summary.AUC <= 0.3, ])

print(paste0("Number of upregulated genes in cap samples: ", n_upregulated))
print(paste0("Number of downregulated genes in cap samples: ", n_downregulated))



```


**Save heatmaps in pdf format**
```{r save-de-genes-in-pdf-format, eval=FALSE}
# Save as R object
very.strong.de.genes.control <- control.markers.wmw[(control.markers.wmw$FDR <= 0.1) & (control.markers.wmw$summary.AUC >= 0.85 |control.markers.wmw$summary.AUC <= 0.15), ]
AUCs.de.very.strong.control <- getMarkerEffects(very.strong.de.genes.control, prefix="AUC")
AUCs.de.very.strong.control <- AUCs.de.very.strong.control[, c("cap50", "cap50_r4h", "cap50_r8h", "cap50_r16h")]

p.de.strong.control.r <- pheatmap(AUCs.de.very.strong.control, breaks=seq(0, 1, length.out=21),
                                color=viridis::viridis(21), cluster_cols = F)
saveRDS(p.de.strong.control.r, file = "./../saved/R_objects/de_genes_plot_control")

# Top de genes
de.genes.control.best <- control.markers.wmw[(control.markers.wmw$Top <= 30) & (control.markers.wmw$FDR <= 0.1) & (control.markers.wmw$summary.AUC >= 0.7 |control.markers.wmw$summary.AUC <= 0.3), ]
AUCs.de.control.best <- getMarkerEffects(de.genes.control.best, prefix="AUC")
AUCs.de.control.best <- AUCs.de.control.best[, c("cap50", "cap50_r4h", "cap50_r8h", "cap50_r16h")]
p.de.control.best <- pheatmap(AUCs.de.control.best, breaks=seq(0, 1, length.out=21),
                                color=viridis::viridis(21), cluster_cols = F)


# Save plot as pdf
## Control
p.de.control <- pheatmap(AUCs.de.control, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21),
                         main = "Marker genes (wmw): Control vs. other samples (FDR <= 0.1 & AUC >= 0.7 | AUC <= 0.3)",
                         fontsize_row = 7, cluster_cols = F)

p.de.strong.control <- pheatmap(AUCs.de.strong.control, breaks=seq(0, 1, length.out=21),
                                color=viridis::viridis(21), main = "Marker genes (wmw): Control vs. other samples (FDR <= 0.1 & AUC >= 0.8 | AUC <= 0.2)",
                                fontsize_row = 7, cluster_cols = F)


pdf("./../saved/figures/de_genes_heatmap_control.pdf", width = 7, height = 45)
p.de.control
dev.off()

pdf("./../saved/figures/de_genes_strong_heatmap_control.pdf", width = 7, height = 15)
p.de.strong.control
dev.off()

## Cap50
## cap50
p.de.cap50 <- pheatmap(AUCs.de.cap50, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21),
                         main = "Marker genes (wmw): cap50 vs. other samples (FDR <= 0.1 & AUC >= 0.7 | AUC <= 0.3)",
                         fontsize_row = 7, cluster_cols = F)

p.de.strong.cap50 <- pheatmap(AUCs.de.strong.cap50, breaks=seq(0, 1, length.out=21),
                                color=viridis::viridis(21), main = "Marker genes (wmw): cap50 vs. other samples (FDR <= 0.1 & AUC >= 0.8 | AUC <= 0.2)",
                                fontsize_row = 7, cluster_cols = F)


pdf("./../saved/figures/de_genes_heatmap_cap50.pdf", width = 7, height = 45)
p.de.cap50
dev.off()

pdf("./../saved/figures/de_genes_strong_heatmap_cap50.pdf", width = 7, height = 15)
p.de.strong.cap50
dev.off()
```


**Markers by cluster**
```{r de-genes-primary-data-cluster-wmw}
# Find markers by sample
markers.primary.clust.wmw <- findMarkers(denoised.sce, 
                                   groups = denoised.sce$cluster,
                                   lfc = log2(1.5),
                                   test = "wilcox")

# Perform marker gene identification for all clusters
for(cluster in unique(denoised.sce$cluster)) {
  # Choose marker for cluster x
  markers.clust.wmw <- markers.primary.clust.wmw[[cluster]]
  # Save marker information to .csv file
  fpath <- paste0("./../saved/tables/marker_genes_cluster_", cluster, "_wmw.csv")
  write.csv(markers.clust.wmw, file = fpath)
  # Subset top 6 genes (ranked by p -value) from each pairwise comparison involving the cluster x
  best.set <- markers.clust.wmw[markers.clust.wmw$Top <= 6,]
  # Extract the area under curves
  AUCs <- getMarkerEffects(best.set, prefix="AUC")
  # Plot results in a heatmap
  pheatmap(AUCs, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21), main = paste("Marker genes (wmw): cluster:", cluster))
}
```


**Markers by cell cycle corrected clusters**
```{r de-genes-primary-data-cluster-wmw-cc}
# Find markers by sample
markers.primary.clust.wmw.cc <- findMarkers(denoised.sce, 
                                   groups = denoised.sce$cluster.cc,
                                   lfc = log2(1.5),
                                   test = "wilcox")

# Perform marker gene identification for all clusters
for(cluster in unique(denoised.sce$cluster.cc)) {
  # Choose marker for cluster x
  markers.clust.wmw <- markers.primary.clust.wmw.cc[[cluster]]
  # Save marker information to .csv file
  fpath <- paste0("./../saved/tables/marker_genes_cluster_", cluster, "cc_wmw.csv")
  write.csv(markers.clust.wmw, file = fpath)
  # Subset top 6 genes (ranked by p -value) from each pairwise comparison involving the cluster x
  best.set <- markers.clust.wmw[markers.clust.wmw$Top <= 6,]
  # Extract the area under curves
  AUCs <- getMarkerEffects(best.set, prefix="AUC")
  # Plot results in a heatmap
  pheatmap(AUCs, breaks=seq(0, 1, length.out=21), color=viridis::viridis(21), main = paste("Marker genes (wmw): cc corrected cluster:", cluster))
}
```


## 2. Ridgeline plots to inspect how the distribution of marker genes differs in the samples of primary and reference datasets
```{r ridge-plots, message=FALSE, warning=FALSE}
# Load de Micheli data
denoised.sce.deMicheli <- readRDS("./../saved/R_objects/denoised_sce_deMicheli")

# Define interesting genes
genes.primary = c("Pax7", "Myf5", "Myod1", "Cdh15", "Tnnt3", "Hes1", "Hes6", "Notch1", "Id1", "Id2", "Id3", "Wnt7b")
genes.deMicheli = c("Pax7", "Myf5", "Myod1", "Myog", "Cdh15", "Tnnt3", "Hes1", "Heyl", "Hes6", "Notch1", "Notch2", "Notch3", "Id1", "Id2", "Id3", "Wnt2", "Wnt10b")

# De Micheli data
sce.deMicheli.f <- denoised.sce.deMicheli[, denoised.sce.deMicheli$cell_type %in% c("d0_MuSCs_and_progenitors", "d5_MuSCs_and_progenitors")]

sce.deMicheli.f$cell_type <- revalue(sce.deMicheli.f$cell_type, c("d0_MuSCs_and_progenitors" = "Day 0", "d5_MuSCs_and_progenitors" = "Day 5"))
seurat.deMicheli <- as.Seurat(sce.deMicheli.f)
Idents(seurat.deMicheli) <- "cell_type"
ridge.deMicheli <- RidgePlot(seurat.deMicheli, features = genes.deMicheli)

# Primary data
seurat.primary <- as.Seurat(denoised.sce)
Idents(seurat.primary) <- "sample"
ridge.primary <- RidgePlot(seurat.primary, features = genes.primary)

ridge.deMicheli
ridge.primary
```

**Save ridge line plots as pdf**
```{r ridge-plots-pdf, eval=FALSE}
pdf("./../saved/figures/ridge_plot_primary.pdf", width = 15, height = 10)
ridge.primary
dev.off()

pdf("./../saved/figures/ridge_plot_deMicheli.pdf", width = 15, height = 10)
ridge.deMicheli
dev.off()
```


